\chapter*{Preface}

This thesis is submitted in partial fulfillment of the requirements for the degree of \gls{phd} at the \gls{ntnu}, Trondheim.

The work presented has been conducted at the \gls{itk}, \gls{ntnu}, as part of the \gls{auf} project.
Professor Kristin Ytterstad Pettersen from the \gls{itk} has been the main supervisor. 
Professor Damiano Varagnolo from the \gls{itk} and Department of Information Engineering, University of Padova, Italy, and doctor Claudio Paliotta from SINTEF Digital and Aker Solutions have been the co-supervisors.
This work was partly supported by the Research Council of Norway through project No. 302435 and the Centres of Excellence funding scheme, project No. 223254.

\section*{Acknowledgments}

First and foremost, I would like to thank my supervisor, Kristin.
Not only she gave me this opportunity to study PhD, but she has also been an excellent supervisor, always positive and optimistic, gently nudging me towards the right direction. 
Furthermore, she possesses an incredible ability to refine my very coarse ideas into something presentable.

I would also like to thank my co-supervisors, Damiano and Claudio, for taking the time off their very busy schedules to guide and help me.
I remember that on multiple occasions, Damiano took a half-day off just to give me detailed feedback on my work, and Claudio has continued to work with me even after leaving SINTEF and getting a new job at Aker Solutions.
Thank you both for that.

I also must not forget to thank my former supervisors from the Czech Technical University in Prague, Tomáš Michálek and Martin Gurtner, as well as the head of the research group, Professor Zdeňek Hurák, for helping me to find the position at NTNU and encouraging me to apply.

The work presented in this thesis is not a work of a single person.
To this end, I would like to thank everyone who contributed to the presented results.
This includes my supervisors, as well as Erlend Basso and Emil Thyri, who helped me with my very first scientific paper, Esteban Restrepo, who visited NTNU for two months in 2022 and with whom we have had a fruitful collaboration ever since, and Erling Lie, whose brilliant Master's project became a conference paper.

Demonstrating that the presented theory actually works typically requires some nice plots and results.
While numerical simulations are nice, there is no substitute for field experiments with very expensive and delicate hardware placed in a hostile environment.
To this end, I would like to thank Professor Jo\~{a}o Tasso de Figueiredo Borges de Sousa and the engineering team at Laborat\'{o}rio de Sistemas e Tecnologias Subaqu\'{a}ticas (LSTS), Faculty of Engineering, University of Porto, and Prof. Martin Ludvigsen, Dr. Antonio Vasilijevic, and the engineering team at the Applied Underwater Robotics Laboratory (AUR-lab), NTNU, for providing access to their underwater vehicles and facilities, and invaluable support before and during the experiments. 
In addition, I wish to thank Tore Mo-Bjørkelund from the AUR-lab and Renato Campos from OceanScan-MST for their help with deploying the control algorithms to the vehicles.

Furthermore, I would like to thank all my friends and colleagues at NTNU for helping me to survive these incredible three years.
Last but not least, I would like to thank my family, my father Josef, my mother Vladislava, and my sister Vladim\'{i}ra, for their support and the relaxing summer and Christmas holidays.
